### 2012-02-21 doc/theses/mikula/proposed_algorithm.tex
Master thesis: In progress.

\begin{define}[Syntax tree]
\label{DEF_syntax_tree}
Syntax tree of XQuery query $Q$ is tuple $T = (V, E, c, \mathcal{P}, o)$ where
\begin{itemize}
\item $V \subset \mathbb{N}$ is a set of nodes, each node representing a particular XQuery construct in query $Q$,
\item $E$ is a set of pairs $(v, w)$ where $v,w \in V$ and for every $a, b \in V, a \neq b: (a, b) \in E$ if and only if a construct represented by $b$ is a direct component of a construct represented by $a$ ($b$ is a child of $a$) in query $Q$,
\item $c: V \rightarrow C$ is function assigning each node with its class from set $C$ of all XQuery language constructs listed in Figures \ref{FIG_syntax_tree_node_types_part_1}, \ref{FIG_syntax_tree_node_types_part_2}, \todo{Tohle je hrozné peklo. Potřebujeme to takto? Nedalo by se to nacpat do 1-2 tabulek/obrázků? Je nutné aby každá položky byla na samostatném řádku. Trochu to vypadá jako umělé natahování diplomky.}
\item $\mathcal{P}$ is a set of functions specifying additional properties of the nodes and distinguishing the nodes of the same classes,
\item and $o : V \rightarrow \mathcal{O}$ is a function specifying an order of children of the nodes, where $\mathcal{O} = \{o_v : E_v \rightarrow \mathbb{N} | v \in V, E_v = \{(v, w) | w \in V, (v,w) \in E\}\}$ is set of functions specifying the children order for each node. For every $v \in V, o(v) = o_v$ so that $o_v(v,w)$ is a sequential number of a construct represented by $w$ amongst constructs represented by children of $v$ in query $Q$.
\end{itemize}
\end{define}

@@

\subsection{Closer Look at the Nodes of the Syntax Tree}
\todo[inline]{Dat figures do appendixu?}
The node classes of the syntax are organized in an is-a hierarchical structure, commonly used in the object oriented programming languages, where an object can be of several types. This hierarchy is shown in tables in Figures \ref{FIG_syntax_tree_node_types_part_1}, \ref{FIG_syntax_tree_node_types_part_2}. The tables are composed of classes in italic and their subclasses. The classes with names in bold represent a common class of a group of subclasses, and these classes cannot be directly used in the syntax tree (abstract classes). The remaining non-bold nodes represent particular constructs of the XQuery language and nodes of these classes can be used in the syntax tree.

@@

\begin{itemize}
\item XML Schema built-in atomic types \todo{link do appendixu}.
\item Types $ElementType$, $AttributeType$, $NodeType$, $TextNodeType$,\linebreak $CommentType$, $ProcessingInstructionType$, $DocumentType$ representing an element, attribute, node, text node, comment, processing instruction, document node.
\item Type representing a node selected by a certain path expression. The path expression is included in this type. Let this type be identified as $PathType$.
\end{itemize}

To capture sequences, we assign each type with its cardinality as proposed in \todo{ref na Schajbalovu diplomku}. Every type can be perceived as a sequence. A type representing one value or one item can be perceived as a sequence of exactly one item. The cardinality expresses one of the following five sequence types.

@@

$PathType$ has always cardinality of zero-or-more, because the result of a path expression is a general sequence and since we do not evaluate the queries, we cannot determine its actual cardinality.

If we cannot determine the type of an expression, we express that using $null$. An example of such case is path expression $/rootElement/(elementA * 2)$. \todo{Nejak dopisat}

@@

\STATE $type = null$
\IF{$is(exprNode, LiteralNode)$}
     \STATE $type =$ XSD atomic type $exprNode.type$, exactly-once cardinality
\ELSIF{$is(exprNode, FunctionCallNode)$}
     \STATE $type = getFunctionReturnType(exprNode.fncName)$
\ELSIF{$is(exprNode, VarRefNode)$}
     \STATE $type = getVariableType(contextVarTypes, exprNode.varName)$
\ELSIF{$is(exprNode, PathExprNode)$}
     \IF{$exprNode$ selects an element set}
         \STATE $type = ElementType$, zero-or-more cardinality, $exprNode$ included
     \ELSIF{$exprNode$ selects an attribute set}
         \STATE $type = AttributeType$, zero-or-more cardinality, $exprNode$ included
     \ENDIF
\ELSIF{$is(exprNode, OperatorNode)$}
     \STATE $type = determineOperatorType(exprNode, contextVarTypes)$
\ELSIF{$is(exprNode, FLWORExprNode)$}
     \STATE $returnClauseNode = exprNode.getChild(ReturnClauseNode)$
     \STATE $type = getType(returnClauseNode)$
\ENDIF

@@

\FORALL{$i \in \{1, \dots, |startingNode.getChildren()|\}$ ordered from the lowest to the highest}
    \STATE $subnode = startingNode.getChild(i)$
    \STATE $analysisOfExpressionTypes(subnode, contextVarTypes)$ \COMMENT{recursion}
    \STATE $newVars = get(subnode, newVars)$    
    \STATE $contextVarTypes = mergeContextVarTypes(contextVarTypes, newVars)$
\ENDFOR
\IF{$is(startingNode, ExprNode)$}
    \STATE $memorizeType(startingNode, determineExpressionType(startingNode))$
\ENDIF
\IF{$is(startingNode, VariableBindingNode)$}
    \STATE $type = null$
    \STATE $typeNode = startingNode.getChild(TypeNode)$
    \IF{$typeNode \neq null$}
        \STATE $type = getTypeTN(typeNode)$
    \ELSE
        \STATE $type = determineExpressionType(startingNode.$ $getChild(BindingSequenceNode).getChild(ExprNode),$ $contextVarTypes)$
    \ENDIF
    \STATE $set(startingNode, newVars, (startingNode.varName, type))$
\ELSIF{$is(startingNode, TupleStreamNode)$}
    \FORALL{$varBindingNode \in startingNode.getChildren()$}
        \STATE $add(startingNode, newVars, get(varBindingNode, newVars))$
    \ENDFOR
\ENDIF
\end{algorithmic}
\end{algorithm}

@@

@@

@@

@@

The first group of XQuery constructs utilized by this step is a group of $distinct-values()$ function and aggregation functions $min()$, $max()$, TODO \todo{Doplnit} as discussed in Section \ref{Negative_uniqueness} and shown in Algorithm \ref{ALG_rejection_of_uniqueness_aggregation_functions}. The algorithm searches for calls of the mentioned functions with a $PathType$ on their input. Elements and attributes selected by those $PathTypes$ are then considered to be not unique, and consequently, they cannot be keys.

Contrary the key discovery, $PathTypes$ utilized in this step may contain predicates. Predicates in a path restrict its result, and therefore, if a set of elements or attributes selected by a path with predicates is not unique, neither is a set of elements or attributes selected by the same path stripped of the predicates. \todo[inline]{Toto asi nebude uplne pravda pre nejake exoticke predikaty. Overit pri implementacii. A: Kdyžtak popište.}

@@

\IF{$is(node, FunctionCallNode)$}
    \IF{$node.fncName$ is one of $distinct-values$, $min$, $max$, TODO}
        \STATE $argument = node.getChild(0)$
        \IF{$getType(argument) == PathType$ using only child and descendant axes}
            \STATE memorize $argument$ as the negative key statement, weight 1
        \ENDIF
    \ENDIF
\ENDIF
\end{algorithmic}
\end{algorithm}

@@

    \FORALL{$bindingNode \in node.getChild(TupleStreamNode).getChildren()$}
        \IF{$is(bindingNode, ForClauseNode)$}
            \STATE $expr = bindingNode.getChild(BindingSequenceNode).$ $getChild(ExprNode)$
            \IF{$getType(expr) == PathExpr$ without predicates, using only child and descendant axes}
                \STATE add $bindingNode.varName$ to $forVars$
            \ENDIF
        \ENDIF
    \ENDFOR

@@

\IF{$is(exprNode, OperatorNode$)}
    \IF{$exprNode.operator == GEN\_EQUALS$}
        \FORALL{$var \in forVars$}
            \IF{$exprNode$ is of a form $var/L = C$ where $L$ is a path without predicates using only child and descendant axes, and getType(C) is a constant}
                \STATE memorize $var/L$ as the negative key statement, weight 0.9
            \ENDIF
        \ENDFOR
    \ELSIF{$exprNode.operator == AND$}
        \STATE processWhere(forVars, exprNode.leftSide)
        \STATE processWhere(forVars, exprNode.rightSide)
    \ENDIF
\ENDIF

@@

\todo[inline]{Doplnit algoritmus comparison with a constant pre predikaty.}

>>>>

\begin{define}[Syntax tree]
\label{DEF_syntax_tree}
Syntax tree of XQuery query $Q$ is tuple $T = (V, E, c, \mathcal{P}, o)$ where
\begin{itemize}
\item $V \subset \mathbb{N}$ is a set of nodes, each node representing a particular XQuery construct in query $Q$,
\item $E$ is a set of pairs $(v, w)$ where $v,w \in V$ and for every $a, b \in V, a \neq b: (a, b) \in E$ if and only if a construct represented by $b$ is a direct component of a construct represented by $a$ ($b$ is a child of $a$) in query $Q$,
\item $c: V \rightarrow C$ is function assigning each node with its class from set $C$ of all XQuery language constructs listed in Figures \ref{FIG_syntax_tree_node_types_part_1}, \ref{FIG_syntax_tree_node_types_part_2},
\item $\mathcal{P}$ is a set of functions specifying additional properties of the nodes and distinguishing the nodes of the same classes,
\item and $o : V \rightarrow \mathcal{O}$ is a function specifying an order of children of the nodes, where $\mathcal{O} = \{o_v : E_v \rightarrow \mathbb{N} | v \in V, E_v = \{(v, w) | w \in V, (v,w) \in E\}\}$ is set of functions specifying the children order for each node. For every $v \in V, o(v) = o_v$ so that $o_v(v,w)$ is a sequential number of a construct represented by $w$ amongst constructs represented by children of $v$ in query $Q$.
\end{itemize}
\end{define}

@@

\subsection{Closer Look at the Nodes of the Syntax Tree}
The node classes of the syntax are organized in an is-a hierarchical structure, commonly used in the object oriented programming languages, where an object can be of several types. This hierarchy is shown in tables in Figures \ref{FIG_syntax_tree_node_types_part_1}, \ref{FIG_syntax_tree_node_types_part_2}. The tables are composed of classes in italic and their subclasses. The classes with names in bold represent a common class of a group of subclasses, and these classes cannot be directly used in the syntax tree (abstract classes). The remaining non-bold nodes represent particular constructs of the XQuery language and nodes of these classes can be used in the syntax tree.

@@

\begin{itemize}
\item XML Schema built-in atomic types \todo{link do appendixu}.
\item Types $ElementType$, $AttributeType$, $NodeType$, $TextNodeType$,\linebreak $CommentType$, $ProcessingInstructionType$, $DocumentType$ representing an element, attribute, node, text node, comment, processing instruction, document node.
\item Type representing a node or a set of nodes selected by a certain path expression. The path expression is included in this type. Let this type be identified as $PathType$.
\item $UnknownType$ representing a type without known details, which does not suit one of the three previous types. An example is XSD type $anyType$.
\end{itemize}

\todo[inline]{Doplnit PathType o boolean for bound, volania zoznam funkcii volanych na tuto path.}

To capture sequences, we assign the first two categories of types (all types except of $PathType$ and $UnknownType$) with its cardinality as proposed in \todo{ref na Schajbalovu diplomku}. Each of those types can be perceived as a sequence. A type representing one value or one item can be perceived as a sequence of exactly one item. The cardinality expresses one of the following five sequence types.

@@

$PathType$ is not assigned with the cardinality since we do not evaluate the queries, and therefore, we cannot determine if a certain XQuery path targets zero, one or more nodes. Alike, $UnknownType$ is neither assigned with the cardinality. Expressions of $UnknownType$ are not utilized it the inference, therefore, their cardinality is not needed.

@@

\STATE $type = null$
\IF{$is(exprNode, LiteralNode)$}
     \STATE $type =$ XSD atomic type $exprNode.type$, exactly-once cardinality
\ELSIF{$is(exprNode, FunctionCallNode)$}
     \STATE $type = getFunctionReturnType(exprNode.fncName)$
\ELSIF{$is(exprNode, VarRefNode)$}
     \STATE $type = getVariableType(contextVarTypes, exprNode.varName)$
\ELSIF{$is(exprNode, PathExprNode)$}
     \STATE $type = createPathType(exprNode)$
\ELSIF{$is(exprNode, OperatorNode)$}
     \STATE $type = determineOperatorType(exprNode, contextVarTypes)$
\ELSIF{$is(exprNode, FLWORExprNode)$}
     \STATE $returnClauseNode = exprNode.getChild(ReturnClauseNode)$
     \STATE $type = createForUnboundType(getType(returnClauseNode))$
\ENDIF

@@

\FORALL{$i \in \{1, \dots, |startingNode.getChildren()|\}$ ordered from the lowest to the highest}
    \STATE $subnode = startingNode.getChild(i)$
    \STATE $analysisOfExpressionTypes(subnode, contextVarTypes)$ \COMMENT{recursion}
    \STATE $newVars = get(subnode, newVars)$    
    \STATE $contextVarTypes = mergeContextVarTypes(contextVarTypes, newVars)$
\ENDFOR
\IF{$is(startingNode, ExprNode)$}
    \STATE $memorizeType(startingNode, determineExpressionType(startingNode))$
\ENDIF
\IF{$is(startingNode, VariableBindingNode)$}
    \STATE $type = null$
    \STATE $typeNode = startingNode.getChild(TypeNode)$
    \IF{$typeNode \neq null$}
        \STATE $type = getTypeTN(typeNode)$
    \ELSE
        \STATE $type = determineExpressionType(startingNode.$ $getChild(BindingSequenceNode).getChild(ExprNode),$ $contextVarTypes)$
        \IF{$is(startingNode, ForClauseNode)$}
            \STATE $type = createForBoundType(type)$
        \ENDIF
    \ENDIF
    \STATE $set(startingNode, newVars, (startingNode.varName, type))$
\ELSIF{$is(startingNode, TupleStreamNode)$}
    \FORALL{$varBindingNode \in startingNode.getChildren()$}
        \STATE $add(startingNode, newVars, get(varBindingNode, newVars))$
    \ENDFOR
\ENDIF
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function createPathType}
\label{ALG_function_create_path_type}
\begin{algorithmic}[1]
\REQUIRE $pathExprNode$: A reference to a PathExprNode to create the PathType from.
\ENSURE The PathType of $pathExprNode$ expression.

\STATE TODO
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function createForBoundType}
\label{ALG_function_create_for_bound_type}
\begin{algorithmic}[1]
\REQUIRE $type$: A type to create the for bound type from.
\ENSURE The for bound type from $type$.

\STATE TODO
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Function createForUnboundType}
\label{ALG_function_create_for_unbound_type}
\begin{algorithmic}[1]
\REQUIRE $type$: A type to create the for unbound type from.
\ENSURE The for unbound type from $type$.

\STATE TODO
\end{algorithmic}
\end{algorithm}

@@

The first group of XQuery constructs utilized by this step is a group of $distinct-values$ function and aggregation functions $min$, $max$, $sum$ as discussed in Section \ref{Negative_uniqueness} and shown in Algorithm \ref{ALG_rejection_of_uniqueness_aggregation_functions}. The algorithm searches for calls of the mentioned functions with a $PathType$ on their input. Elements and attributes selected by those $PathTypes$ are then considered to be not unique, and consequently, they cannot be keys.

Contrary the key discovery, $PathTypes$ utilized in this step may contain predicates in their lasts steps. Predicates in a path's last step restrict its result, and therefore, if a set of elements or attributes selected by a path with predicates is not unique, neither is a set of elements or attributes selected by the same path stripped of the predicates.

@@

\IF{$is(node, FunctionCallNode)$}
    \IF{$node.fncName$ is one of $distinct-values$, $min$, $max$, $sum$}
        \STATE $argument = node.getChild(0)$
        \IF{$getType(argument) == PathType$ using only child and descendant axes, without predicates except the last step}
            \STATE memorize $argument$ as the negative key statement, weight 1
        \ENDIF
    \ENDIF
\ENDIF
\end{algorithmic}
\end{algorithm}

@@

    \FORALL{$bindingNode \in node.getChild(TupleStreamNode).getChildren()$}
        \IF{$is(bindingNode, ForClauseNode)$}
            \STATE $expr = bindingNode.getChild(BindingSequenceNode).$ $getChild(ExprNode)$
            \IF{$getType(expr) == PathExpr$ without predicates except the last step, using only child and descendant axes}
                \STATE add $bindingNode.varName$ to $forVars$
            \ENDIF
        \ENDIF
    \ENDFOR

@@

\IF{$is(exprNode, OperatorNode$)}
    \IF{$exprNode.operator == GEN\_EQUALS$}
        \FORALL{$var \in forVars$}
            \IF{$exprNode$ is of a form $var/L = C$ where $L$ is a path without predicates, except the last step, using only child and descendant axes, and C is a literal constant}
                \STATE memorize $var/L$ as the negative key statement, weight 0.9
            \ENDIF
        \ENDFOR
    \ELSIF{$exprNode.operator == AND$}
        \STATE processWhere(forVars, exprNode.leftSide)
        \STATE processWhere(forVars, exprNode.rightSide)
    \ENDIF
\ENDIF

@@
